A Framework for Equivalence of Cast Calculi and a new Cast Representation

### 2020-10-05 paper.tex
edits to sec 5

\begin{document}
%
\title{A Framework for Equivalence of Cast Calculi and a new Cast Representation}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Kuang-Chen Lu\inst{1} \and
Jeremy G. Siek\inst{2} \and
Andre Kuhlenschmidt\inst{2}
}
%
\authorrunning{F. Author et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Brown University, Providence RI 02912, USA \and
Indiana University, Bloomington IN 47405, USA}
% \institute{Princeton University, Princeton NJ 08544, USA \and
% Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany
% \email{lncs@springer.com}\\
% \url{http://www.springer.com/gp/computer-science/lncs} \and
% ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
% \email{\{abc,lncs\}@uni-heidelberg.de}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Gradual typing aims at combining dynamic typing and static typing in one 
language to provide the advantages of both typing disciplines. Implementing 
gradual typing with reasonable overhead, however, is not trivial. Casts might 
accumulate in higher-order values and continuations, thus change the overall 
space complexity. Several space-efficient cast representations have been 
proposed, notably coercions in normal forms and threesomes. But they are
presented in different dynamic semantics, which make it difficult to compare
the key differences between the cast representations. We present a abstrat
data type, namely Cast ADT, that captures interface functions a cast representations
typically includes, and an abstract machine parameterized over Cast ADTs.
The Cast ADT brings another benefit: space-efficient cast representations 
shall be proved not changing the 
overall behavior of language implementations except saving space.
With our generic abstract machine, these proofs can be done by proving properties
of a specific Cast ADT, which is simpler than going through a bisimulation proof 
between a space-efficient dynamic semantics and the definitional dynamic semantics.
We have successfully applied our new approach of proving the correctness of cast 
representations to several known cast representations and believe this framework
is also applicable to other cast representations.
\todo[inline]{JS: cite which other cast representations}
We also present a new cast representation, hypercoercions, that is more friendly
to mechanized meta-theory and hypothetically consume less memory than coercions
in normal forms.

@@

\todo[inline]{KC: I changed some wording in the following definition:
"a family"; translate a cast ...}
\begin{definition}[Cast Abstract Data Type (Cast ADT)]
	\label{def:cast-rep}
	A cast abstract data type is a family, namely $Cast$, indexed by two types, 
	with four operators and two properties of operators:
\begin{itemize}
	\item $\id{T}=c$: create an identity cast
	\item $\compose{c}{c}=c$: compose casts
	\item $\translate{\ccast{S}{l}{T}}=c$: translate a cast in source 
		program to a $\textit{Cast}_{S,T}$
	\item $\denote{c}(v)=r$: apply a cast to a value
	\item If $ \valuetyping{v}{T} $ then $\denote{\id{T}}(v)=v$
	\item If $ \valuetyping{v}{T_1} $ and
		$ \casttyping{c}{T_1}{T_2} $ and
		$ \casttyping{d}{T_2}{T_3} $ 
		then 
		$ \denote{\compose{c}{d}}(v) = (\denote{c}(v)\mbind\denote{d})$
\end{itemize}
We use the syntax \casttyping{c}{S}{T} to mean $c$ is in the set 
$\textit{Cast}_{S,T}$.
\end{definition}

@@

Subsection~\ref{sec:LazyD-Hypercoercions} implements the \CastADT\ with the \LD\ 
hypercoercions and proves that \LD\ hypercoercions is a Lazy D Cast ADT. 
Subsection~\ref{sec:LazyUD-Hypercoercions} does the \LUD\ counterparts.
Finally, we argue in Subsection~\ref{sec:hypercoercions-compact} that 
hypercoercions would consume less space than coercions in 
normal form and hence are more suitable for implementation.

@@

\fbox{$\goodgap{g}$}
\begin{gather*}
\infr{}{\goodgap{\gnone}} \quad
\infr{\shallowlyconsistent{P}{Q}}{\goodgap{\gsome{P}{l}{Q}}}
\end{gather*}

@@

\fbox{$\snd{g}=g$}
\[
\begin{array}{rclr}
\figequalto{\snd{\gnone}}{\gnone}{}
\figequalto{\snd{\gsome{\Pprod{S_1}{T_1}}{l}{\Pprod{S_2}{T_2}}}}{
	\gsome{T_1}{l}{T_2}
}{}
\end{array}
\]

\figref{fig:LazyD-Hypercoercions-composition} defines the functions 
\compose{c}{c} that compose \LD\ hypercoercions. We use these functions to 
implement the \CastADT.

\todo[inline]{KC: (About the following paragraph) Should we mention KC and Deyaa as citation?
(something like in private communication with KC and Deyaa)}

The \compose{c}{c} operator is the key to compressing coercions and maintaining 
space-efficiency. The composition operator for Coercions in Normal Form 
requires the target type of the first coercion to match the source type of the 
second. This requirement, however, imposes a considerable difficulty in 
termination proof of the composition of \LD\ Coercions in Normal Form, reported 
by Kuang-Chen Lu and Deyaaeldeen Almahallawi. So it is useful to relax this 
restriction for the \LD\ tracking strategy.
%
We shall need optional source cast, written $g$, that range over
$\gnone$ or $\gsome{S}{l}{T}$, to maintain the gap between the hypercoercions 
to be composed. The source cast is mandatory when the target
type of the first hypercoercion does not match the source type of the second. 
Then we write the composition of hypercoercions as \composegap{c_1}{g}{c_2}.
%  
When both $c_1$ and $c_2$ are \hcci, their composition is also \hcci. 
%
When the first hypercoercion is the \hcci, the output must start with a 
projection, in which case we invoke the helper function \textit{mkProj}. In 
\textit{mkProj}, if the head of $c_2$ is a projection, the result is $c_2$.  
Otherwise, we need a label to build the projection. Since the head of $c_2$ is 
the no-op, its source type must be a pretype. Thus we know $g$ must be a 
source cast and we use it to construct the projection.
%
Going back to $\composegap{c}{g}{c}$, when the second hypercoercion is the 
\hcci, the output must end with an injection, in which case we invoke the 
helper function \textit{mkInj}.
%
Finally, when both inputs are of the form \hccc{h}{m}{t}, we compose the middle 
with an updated gap.
%
The second and third cases of \mkgap{t}{g}{h} is interesting, where we choose 
the label in projection to be responsible for potential inconsistency.

Another interesting operator is the middle composition \composegap{m}{g}{m}. 
%
When the first middle is a failure (\hcmfail{l}), the result is the second
middle. In all the remaining cases, the first middle is not a failure.
%
When the gap is a cast between shallowly inconsistent types (i.e. having 
different top-most constructors). the output is a failure.
%
When the gap is good but the second middle is a failure, the output is the same 
failure. A gap $g$ is good if and only if there is not $P$ and $l$ 
and $Q$ such that $g = \gsome{P}{l}{Q}$ and \notshallowlyconsistent{P}{Q}.
%
The remaining cases are straightforward recursion.

Agda is able to understand that \composegap{c}{g}{c} is terminating. The reason 
might be that in \composegap{c}{g}{c} both casts become smaller in recursive 
calls, although their position is swapped in the function case. Readers who
hate non-structural recursion might add a Boolean parameter to indicate the order
of composition.

\begin{figure}[tp]
	\fbox{$\translate{\ccast{S}{l}{T}} = c$} Translation of \LD\ hypercoercions
	\[
	\begin{array}{rclr}
	\figequalto{\translate{\ccast{S}{l}{T}}}{\composegap{id(S)}{(\gsome{S}{l}{T})}{id(T)}}{}
	\end{array}
	\]
	\caption{Translatin of \LD\ Hypercoercions}
	\label{fig:LazyD-Hypercoercions-translation}
\end{figure}

The definition of translation \translate{\ccast{S}{l}{T}} is straightforward 
(\figref{fig:LazyD-Hypercoercions-translation}).

@@

Fig.~\ref{fig:LazyD-Hypercoercions-application} defines the cast application 
function for \LD\ hypercoercions.
Applying the identity cast for the dynamic type succeeds immediately. 
Otherwise, we apply head, middle, and tail in sequence. 
%
In projection case of head application, a new middle is constructed and applied 
to the injected value.
%
Middle applications raises an error if the middle is a failure and extend value 
proxies otherwise.

>>>>

\begin{document}
%
\title{A Framework for Equivalence of Cast Calculi and a new Cast Representation}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Kuang-Chen Lu\inst{1} \and
Jeremy G. Siek\inst{2} \and
Andre Kuhlenschmidt\inst{2}
}
%
\authorrunning{F. Author et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Brown University, Providence RI 02912, USA \and
Indiana University, Bloomington IN 47405, USA}
% \institute{Princeton University, Princeton NJ 08544, USA \and
% Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany
% \email{lncs@springer.com}\\
% \url{http://www.springer.com/gp/computer-science/lncs} \and
% ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
% \email{\{abc,lncs\}@uni-heidelberg.de}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Gradual typing aims at combining dynamic typing and static typing in one 
language to provide the advantages of both typing disciplines. Implementing 
gradual typing with reasonable overhead, however, is not trivial. Casts might 
accumulate in higher-order values and continuations, thus change the overall 
space complexity. Several space-efficient cast representations have been 
proposed, notably coercions in normal forms and threesomes. But they are
presented in different dynamic semantics, which make it difficult to compare
the key differences between the cast representations. We present a abstrat
data type, namely Cast ADT, that captures interface functions a cast representations
typically includes, and an abstract machine parameterized over Cast ADTs.
The Cast ADT brings another benefit: space-efficient cast representations 
shall be proved not changing the 
overall behavior of language implementations except saving space.
With our generic abstract machine, these proofs can be done by proving properties
of a specific Cast ADT, which is simpler than going through a bisimulation proof 
between a space-efficient dynamic semantics and the definitional dynamic semantics.
We have successfully applied our new approach of proving the correctness of cast 
representations to several known cast representations and believe this framework
is also applicable to other cast representations.
\todo[inline]{JS: cite which other cast representations}
We also present a new cast representation, hypercoercions, that is more friendly
to mechanized meta-theory and consumes less memory than coercions
in normal forms.

@@

\begin{definition}[Cast Abstract Data Type (Cast ADT)]
	\label{def:cast-rep}
	A cast abstract data type is a family, namely $Cast$, indexed by two types, 
	with four operators and two properties of operators:
\begin{itemize}
	\item $\id{T}$: create an identity cast
	\item $\compose{c}{c}$: compose two casts
	\item $\translate{\ccast{S}{l}{T}}$: translate a cast in the source 
		program to a $\textit{Cast}_{S,T}$
	\item $\denote{c}(v)$: apply a cast to a value
	\item If $ \valuetyping{v}{T} $ then $\denote{\id{T}}(v)=v$
	\item If $ \valuetyping{v}{T_1} $ and
		$ \casttyping{c}{T_1}{T_2} $ and
		$ \casttyping{d}{T_2}{T_3} $ 
		then 
		$ \denote{\compose{c}{d}}(v) = (\denote{c}(v)\mbind\denote{d})$
\end{itemize}
We use the syntax \casttyping{c}{S}{T} to mean $c$ is in the set 
$\textit{Cast}_{S,T}$.
\end{definition}

@@

Subsection~\ref{sec:LazyD-Hypercoercions} implements the \CastADT{}
with the \LD{} hypercoercions and proves that \LD{} hypercoercions is
a Lazy D Cast ADT.  Subsection~\ref{sec:LazyUD-Hypercoercions} does
the same for \LUD{}.  Finally, we argue in
Subsection~\ref{sec:hypercoercions-compact} that hypercoercions
consume less space than coercions in normal form and hence are more
suitable for implementation.

@@

\begin{figure}[tp]
\fbox{$\goodgap{g}$}
\begin{gather*}
\infr{}{\goodgap{\gnone}} \quad
\infr{\shallowlyconsistent{P}{Q}}{\goodgap{\gsome{P}{l}{Q}}}
\end{gather*}

@@

\fbox{$\snd{g}=g$}
\[
\begin{array}{rclr}
\figequalto{\snd{\gnone}}{\gnone}{}
\figequalto{\snd{\gsome{\Pprod{S_1}{T_1}}{l}{\Pprod{S_2}{T_2}}}}{
	\gsome{T_1}{l}{T_2}
}{}
\end{array}
\]
\caption{Auxilliary predicates and functions for hypercoercion composition.}
\label{fig:aux-hypercoercion-composition}
\end{figure}

\figref{fig:LazyD-Hypercoercions-composition} defines the composition
operator \compose{c}{c} for \LD\ hypercoercions and
\figref{fig:aux-hypercoercion-composition} defines some auxilliary
predicates and functions.

\todo[inline]{KC: (About the following paragraph) Should we mention KC and Deyaa as citation?
  (something like in private communication with KC and Deyaa)}
\todo[inline]{JS: Yes}

The \compose{c}{c} operator is the key to compressing coercions and
maintaining space-efficiency. The composition operator for Coercions
in Normal Form requires the target type of the first coercion to match
the source type of the second. This requirement, however, imposes
considerable difficulty in the termination proof of the composition
for \LD\ Coercions in Normal Form, reported by Kuang-Chen Lu and
Deyaaeldeen Almahallawi. So it is useful to relax this restriction for
the \LD\ tracking strategy.
%
We shall need an optional source cast, written $g$, that ranges over
$\gnone$ or $\gsome{S}{l}{T}$, to maintain the gap between the two
hypercoercions to be composed. The source cast is mandatory when the
target type of the first hypercoercion does not match the source type
of the second.  Then we write the composition of hypercoercions as
\composegap{c_1}{g}{c_2}.
%  
When both $c_1$ and $c_2$ are \hcci, their composition is also \hcci. 
%
When the first hypercoercion is the \hcci, the output must start with
a projection, in which case we invoke the helper function
\textit{mkProj}. In \textit{mkProj}, if the head of $c_2$ is a
projection, the result is $c_2$.  Otherwise, we need a label to build
the projection. Since the head of $c_2$ is a no-op, its source type
must be a pre-type. Thus we know $g$ must be a source cast and we use
it to construct the projection.
%
Going back to $\composegap{c}{g}{c}$, when the second hypercoercion is the 
\hcci, the output must end with an injection, in which case we invoke the 
helper function \textit{mkInj}.
%
Finally, when both inputs are of the form \hccc{h}{m}{t}, we compose the middle 
with an updated gap.
%
The second and third cases of \mkgap{t}{g}{h} is interesting, where we choose 
the label in projection to be responsible for potential inconsistency.

Another interesting operator is the middle composition \composegap{m}{g}{m}. 
%
When the first middle is a failure (\hcmfail{l}), the result is the second
middle. In all the remaining cases, the first middle is not a failure.
%
When the gap is a cast between shallowly inconsistent types (i.e. having 
different top-most constructors) the output is a failure.
%
When the gap is good but the second middle is a failure, the output is the same 
failure. A gap $g$ is good if and only if there is not $P$ and $l$ 
and $Q$ such that $g = \gsome{P}{l}{Q}$ and \notshallowlyconsistent{P}{Q}.
%
The remaining cases are straightforward recursion.

Agda is able to understand that \composegap{c}{g}{c} is
terminating. The reason might be that in \composegap{c}{g}{c} both
casts become smaller in recursive calls, although their position is
swapped in the function case. Readers who dislike non-structural
recursion could add a Boolean parameter to indicate the direction of
composition.

\begin{figure}[tp]
	\fbox{$\translate{\ccast{S}{l}{T}} = c$} Translation of \LD\ hypercoercions
	\[
	\begin{array}{rclr}
	\figequalto{\translate{\ccast{S}{l}{T}}}{\composegap{id(S)}{(\gsome{S}{l}{T})}{id(T)}}{}
	\end{array}
	\]
	\caption{Translation of \LD\ Hypercoercions}
	\label{fig:LazyD-Hypercoercions-translation}
\end{figure}

The definition of the translation function \translate{\ccast{S}{l}{T}}
is straightforward (\figref{fig:LazyD-Hypercoercions-translation}).

@@

Fig.~\ref{fig:LazyD-Hypercoercions-application} defines the cast application 
function for \LD\ hypercoercions.
Applying the identity cast for the dynamic type succeeds immediately. 
Otherwise, we apply head, middle, and tail in sequence. 
%
In the projection case of applying a head coercion, a new middle is
constructed and applied to the injected value.
%
Middle applications raise an error if the middle is a failure,
otherwise they compose their coercions with those in the value.
